---
title: "Forest Impact Simulator"
author: "Forest Carbon Analysis Tool"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(tidyverse)
library(ggplot2)
library(jsonlite)
library(readr)
library(dplyr)
library(plotly)
```

# Forest Impact Simulator

This notebook simulates the environmental impact of forest planting or clear-cutting activities. It calculates carbon sequestration/loss, biodiversity impact, and other environmental metrics.

## 1. Data Input

### Option A: Upload CSV File

If you have a CSV file with forest data, upload it here. The file should contain columns for land identification, area, tree types, and carbon data.

```{r data-input, echo=FALSE, message=FALSE}
# Check if sample data file exists
sample_file <- "sample-forest-planting-data.csv"
if (file.exists(sample_file)) {
  cat("**Sample data file found!** Loading structured CSV data...\n\n")
  
  # Read the sample data file
  lines <- readLines(sample_file)
  
  # Parse the structured data
  metadata_start <- which(lines == "METADATA") + 1
  trees_start <- which(lines == "SELECTED TREES") + 1
  env_start <- which(lines == "ENVIRONMENTAL DATA") + 1
  impact_start <- which(lines == "IMPACT RESULTS") + 1
  planting_start <- which(lines == "PLANTING DATA") + 1
  
  # Extract metadata
  metadata_lines <- lines[metadata_start:(trees_start-2)]
  metadata <- data.frame(
    Parameter = sapply(strsplit(metadata_lines, ","), `[`, 1),
    Value = sapply(strsplit(metadata_lines, ","), function(x) paste(x[-1], collapse = ","))
  )
  
  # Extract tree data
  tree_lines <- lines[trees_start:(env_start-2)]
  tree_header <- strsplit(tree_lines[1], ",")[[1]]
  tree_data <- data.frame(
    Name = sapply(strsplit(tree_lines[-1], ","), `[`, 1),
    Scientific_Name = sapply(strsplit(tree_lines[-1], ","), `[`, 2),
    Carbon_Sequestration = as.numeric(sapply(strsplit(tree_lines[-1], ","), `[`, 3)),
    Percentage = sapply(strsplit(tree_lines[-1], ","), `[`, 4)
  )
  
  # Extract environmental data
  env_lines <- lines[env_start:(impact_start-2)]
  env_data <- data.frame(
    Metric = sapply(strsplit(env_lines[-1], ","), `[`, 1),
    Value = sapply(strsplit(env_lines[-1], ","), function(x) paste(x[-1], collapse = ","))
  )
  
  # Extract impact results
  impact_lines <- lines[impact_start:(planting_start-2)]
  impact_data <- data.frame(
    Metric = sapply(strsplit(impact_lines[-1], ","), `[`, 1),
    Value = sapply(strsplit(impact_lines[-1], ","), function(x) paste(x[-1], collapse = ","))
  )
  
  # Extract planting data
  planting_lines <- lines[planting_start:length(lines)]
  planting_data <- data.frame(
    Metric = sapply(strsplit(planting_lines[-1], ","), `[`, 1),
    Value = sapply(strsplit(planting_lines[-1], ","), function(x) paste(x[-1], collapse = ","))
  )
  
  # Display loaded data
  cat("**Metadata:**\n")
  print(metadata)
  
  cat("\n**Tree Data:**\n")
  print(tree_data)
  
  cat("\n**Environmental Data:**\n")
  print(env_data)
  
  cat("\n**Impact Results:**\n")
  print(impact_data)
  
  cat("\n**Planting Data:**\n")
  print(planting_data)
  
  # Set flag to indicate we have structured data
  has_structured_data <- TRUE
  
} else {
  cat("**No sample data file found.** Please provide your own data or use manual input below.\n\n")
  has_structured_data <- FALSE
}
```

### Option B: Manual Data Input

If you don't have a CSV file, you can manually input your forest data here:

```{r manual-input, echo=FALSE}
# Manual data input section
cat("## Manual Data Input\n\n")
cat("Please modify the following data frames with your forest data:\n\n")

# Create sample data structure for manual input
land_data <- data.frame(
  land_id = c("Plot_001", "Plot_002", "Plot_003"),
  area_ha = c(0.1, 0.15, 0.2),
  tree_type = c("Oak", "Pine", "Mixed"),
  carbon_per_ha = c(110, 125, 118),
  biodiversity_score = c(4.5, 4.2, 4.8),
  resilience_score = c(4.3, 4.0, 4.6),
  stringsAsFactors = FALSE
)

cat("**Land Plot Data:**\n")
print(land_data)

# Location and simulation parameters
location_data <- data.frame(
  latitude = 50.56956736416948,
  longitude = 28.159344338632376,
  simulation_years = 50
)

cat("\n**Location and Simulation Parameters:**\n")
print(location_data)
```

## 2. Mode Selection

Select the simulation mode:

```{r mode-selection, echo=FALSE}
# Mode selection
mode <- "planting"  # Change to "clear-cutting" for deforestation simulation

cat(paste("**Selected Mode:**", toupper(mode), "\n\n"))

if (mode == "planting") {
  cat("🌱 **Planting Mode**: Calculating positive carbon sequestration from new forest growth\n")
} else {
  cat("🌲 **Clear-cutting Mode**: Calculating carbon loss from forest removal\n")
}
```

## 3. Data Processing and Calculations

```{r calculations, echo=TRUE}
# Process the data based on mode
if (exists("has_structured_data") && has_structured_data) {
  # Use structured data from CSV file
  current_data <- tree_data
  area_ha <- as.numeric(planting_data$Value[planting_data$Metric == "Area (hectares)"])
  simulation_years <- as.numeric(metadata$Value[metadata$Parameter == "Simulation Years"])
  latitude <- as.numeric(metadata$Value[metadata$Parameter == "Latitude"])
  longitude <- as.numeric(metadata$Value[metadata$Parameter == "Longitude"])
  
  # Extract environmental values
  soil_carbon <- as.numeric(env_data$Value[env_data$Metric == "Soil Carbon (g/kg)"])
  soil_ph <- as.numeric(env_data$Value[env_data$Metric == "Soil pH"])
  temperature <- as.numeric(env_data$Value[env_data$Metric == "Temperature (°C)"])
  precipitation <- as.numeric(env_data$Value[env_data$Metric == "Precipitation (mm)"])
  
  # Create land data from existing structure
  land_data <- data.frame(
    land_id = paste0("Plot_", 1:nrow(current_data)),
    area_ha = area_ha / nrow(current_data),
    tree_type = current_data$Name,
    carbon_per_ha = current_data$Carbon_Sequestration,
    biodiversity_score = rep(4.5, nrow(current_data)),
    resilience_score = rep(4.5, nrow(current_data)),
    stringsAsFactors = FALSE
  )
  
  location_data <- data.frame(
    latitude = latitude,
    longitude = longitude,
    simulation_years = simulation_years
  )
  
  # Create structured environmental data
  env_data_structured <- data.frame(
    soil_carbon = ifelse(is.na(soil_carbon), NA, soil_carbon),
    soil_ph = ifelse(is.na(soil_ph), NA, soil_ph),
    temperature = ifelse(is.na(temperature), 22.4, temperature),
    precipitation = ifelse(is.na(precipitation), NA, precipitation)
  )
  
} else {
  # Use manual data
  simulation_years <- location_data$simulation_years
  env_data_structured <- data.frame(
    soil_carbon = NA,
    soil_ph = NA,
    temperature = 22.4,
    precipitation = NA
  )
}

# Calculate CO2 impact based on mode
if (mode == "planting") {
  land_data$co2_impact_tons <- land_data$area_ha * land_data$carbon_per_ha * simulation_years
  mode_multiplier <- 1
  impact_type <- "Carbon Sequestration"
} else {
  land_data$co2_impact_tons <- -(land_data$area_ha * land_data$carbon_per_ha * simulation_years)
  mode_multiplier <- -1
  impact_type <- "Carbon Loss"
}

# Calculate totals and averages
total_carbon <- sum(land_data$co2_impact_tons)
average_biodiversity <- mean(land_data$biodiversity_score)
average_resilience <- mean(land_data$resilience_score)
total_area <- sum(land_data$area_ha)
total_trees <- round(total_area * 625)  # Assuming 625 trees per hectare

# Display results
cat("## Calculation Results\n\n")
cat("**Land Plot Calculations:**\n")
print(land_data)

cat(paste("\n**Summary Statistics:**\n"))
cat(paste("- Total", impact_type, ":", round(total_carbon, 2), "kg CO2\n"))
cat(paste("- Total Area:", round(total_area, 3), "hectares\n"))
cat(paste("- Average Biodiversity Score:", round(average_biodiversity, 2), "/5\n"))
cat(paste("- Average Resilience Score:", round(average_resilience, 2), "/5\n"))
cat(paste("- Total Trees:", total_trees, "\n"))
```

## 4. Data Visualization

### CO2 Impact by Land Plot

```{r co2-bar-chart, fig.width=10, fig.height=6}
# Create bar chart for CO2 impact per land plot
co2_plot <- ggplot(land_data, aes(x = land_id, y = co2_impact_tons, fill = tree_type)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = paste(round(co2_impact_tons, 1), "kg CO2")), 
            vjust = -0.5, size = 3.5) +
  labs(
    title = paste("CO2 Impact by Land Plot -", impact_type),
    subtitle = paste("Total:", round(total_carbon, 2), "kg CO2 over", simulation_years, "years"),
    x = "Land Plot ID",
    y = paste("CO2", ifelse(mode == "planting", "Sequestration", "Loss"), "(kg)"),
    fill = "Tree Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")

print(co2_plot)
```

### Environmental Metrics Comparison

```{r metrics-comparison, fig.width=10, fig.height=6}
# Create comparison chart for environmental metrics
metrics_data <- land_data %>%
  select(land_id, biodiversity_score, resilience_score) %>%
  pivot_longer(cols = c(biodiversity_score, resilience_score), 
               names_to = "metric", values_to = "score") %>%
  mutate(metric = case_when(
    metric == "biodiversity_score" ~ "Biodiversity Impact",
    metric == "resilience_score" ~ "Forest Resilience"
  ))

metrics_plot <- ggplot(metrics_data, aes(x = land_id, y = score, fill = metric)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_text(aes(label = paste(round(score, 1), "/5")), 
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3.5) +
  labs(
    title = "Environmental Metrics by Land Plot",
    x = "Land Plot ID",
    y = "Score (out of 5)",
    fill = "Metric"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ylim(0, 5)

print(metrics_plot)
```

### Area Distribution

```{r area-pie-chart, fig.width=8, fig.height=8}
# Create pie chart for area distribution
area_plot <- ggplot(land_data, aes(x = "", y = area_ha, fill = tree_type)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste(tree_type, "\n", round(area_ha, 3), "ha")), 
            position = position_stack(vjust = 0.5), size = 4) +
  labs(
    title = "Area Distribution by Tree Type",
    fill = "Tree Type"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "none"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")

print(area_plot)
```

## 5. Output Generation

### Save Processed Data as CSV

```{r save-csv}
# Create output directory if it doesn't exist
if (!dir.exists("output")) {
  dir.create("output")
}

# Save processed land data
output_csv <- "output/processed_forest_data.csv"
write.csv(land_data, output_csv, row.names = FALSE)

cat(paste("✅ Processed data saved to:", output_csv, "\n\n"))
```

### Generate JSON Output (Web App Compatible)

```{r generate-json}
# Create JSON structure matching web app format
timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%S.000Z")

# Calculate additional metrics
water_retention <- 94  # Default value, could be calculated from environmental data
air_quality_improvement <- 95  # Default value, could be calculated from environmental data
annual_carbon <- total_carbon / simulation_years

# Create the JSON structure
json_output <- list(
  metadata = list(
    timestamp = timestamp,
    simulatorVersion = "1.0.0",
    location = list(
      latitude = location_data$latitude,
      longitude = location_data$longitude,
      type = "Point"
    ),
    simulation = list(
      years = simulation_years,
      mode = mode,
      totalPlots = nrow(land_data)
    )
  ),
  environmentalData = list(
    soil = list(
      carbon = env_data_structured$soil_carbon,
      pH = env_data_structured$soil_ph
    ),
    climate = list(
      temperature = env_data_structured$temperature,
      precipitation = env_data_structured$precipitation
    )
  ),
  impactResults = list(
    carbonSequestration = ifelse(mode == "planting", annual_carbon, -annual_carbon),
    biodiversityImpact = average_biodiversity,
    forestResilience = average_resilience,
    waterRetention = water_retention,
    airQualityImprovement = air_quality_improvement,
    totalCarbon = total_carbon,
    averageBiodiversity = average_biodiversity,
    averageResilience = average_resilience
  ),
  plantingData = list(
    area = total_area,
    totalTrees = total_trees,
    spacing = 4,
    density = 625,
    timeline = simulation_years,
    plots = land_data
  )
)

# Save JSON file
output_json <- "output/forest_impact_results.json"
write_json(json_output, output_json, pretty = TRUE, auto_unbox = TRUE)

cat(paste("✅ JSON results saved to:", output_json, "\n\n"))

# Display summary
cat("## Final Summary\n\n")
cat("**Simulation Results:**\n")
cat(paste("- Mode:", toupper(mode), "\n"))
cat(paste("- Total", impact_type, ":", round(total_carbon, 2), "kg CO2\n"))
cat(paste("- Annual", impact_type, ":", round(annual_carbon, 2), "kg CO2/year\n"))
cat(paste("- Simulation Period:", simulation_years, "years\n"))
cat(paste("- Number of Plots:", nrow(land_data), "\n"))
cat(paste("- Total Area:", round(total_area, 3), "hectares\n"))
cat(paste("- Average Biodiversity:", round(average_biodiversity, 2), "/5\n"))
cat(paste("- Average Resilience:", round(average_resilience, 2), "/5\n"))
```

## 6. Additional Analysis (Optional)

### Yearly Carbon Accumulation Projection

```{r yearly-projection, fig.width=12, fig.height=6, eval=FALSE}
# Optional: Create yearly carbon accumulation projection
if (simulation_years > 1 && mode == "planting") {
  yearly_data <- expand.grid(
    year = 1:simulation_years,
    land_id = land_data$land_id,
    tree_type = land_data$tree_type
  ) %>%
    left_join(land_data, by = c("land_id", "tree_type")) %>%
    mutate(
      # Assume linear growth for simplicity (could be more complex)
      cumulative_carbon = (area_ha * carbon_per_ha * year) / simulation_years
    )
  
  yearly_plot <- ggplot(yearly_data, aes(x = year, y = cumulative_carbon, color = tree_type)) +
    geom_line(size = 1.2) +
    geom_point(size = 2) +
    labs(
      title = "Cumulative Carbon Sequestration Over Time",
      x = "Year",
      y = "Cumulative Carbon (kg CO2)",
      color = "Tree Type"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom"
    ) +
    scale_color_brewer(type = "qual", palette = "Set2")
  
  print(yearly_plot)
}
```

---

## Instructions for Use

1. **Data Input**: 
   - **Structured CSV**: Upload a CSV file with the exact format shown in `sample-forest-planting-data.csv` containing METADATA, SELECTED TREES, ENVIRONMENTAL DATA, IMPACT RESULTS, and PLANTING DATA sections
   - **Manual Input**: Modify the manual input section with your data if no CSV file is available

2. **Mode Selection**: Change the `mode` variable to either "planting" or "clear-cutting" based on your simulation needs.

3. **Run Analysis**: Execute all code chunks sequentially to perform calculations and generate visualizations.

4. **Output Files**: Check the `output/` directory for:
   - `processed_forest_data.csv`: Processed data in CSV format
   - `forest_impact_results.json`: Results in JSON format compatible with the web app

5. **CSV Format**: The notebook supports the structured CSV format with sections:
   - **METADATA**: Timestamp, simulator version, simulation years, location coordinates
   - **SELECTED TREES**: Tree name, scientific name, carbon sequestration rate, percentage
   - **ENVIRONMENTAL DATA**: Soil carbon, pH, temperature, precipitation
   - **IMPACT RESULTS**: Annual carbon sequestration, total carbon, biodiversity, resilience
   - **PLANTING DATA**: Area, total trees, spacing, density

6. **Customization**: Modify the environmental metrics, tree types, or calculation methods as needed for your specific use case.

## Notes

- Carbon calculations assume linear accumulation over the simulation period
- Biodiversity and resilience scores are on a 1-5 scale
- Water retention and air quality improvement are set to default values (can be customized)
- The JSON output structure matches the Forest Impact Simulator web app format
- All calculations are based on the input data and selected simulation mode

---

*Generated by Forest Impact Simulator R Notebook - Version 1.0.0*
